Accelerated and interpretable oblique random survival forests

### 2022-07-27 paper/arxiv/main.Rnw
pure time benchmark plus arxiv

\documentclass[twoside,11pt]{article}

% Any additional packages needed should be included after jmlr2e.
% Note that jmlr2e.sty includes epsfig, amssymb, natbib and graphicx,
% and defines many common macros, such as 'proof' and 'example'.
%
% It also sets the bibliographystyle to plainnat; for more information on
% natbib citation styles, see the natbib documentation, a copy of which
% is archived at http://www.jmlr.org/format/natbib.pdf

% Available options for package jmlr2e are:
%
%   - abbrvbib : use abbrvnat for the bibliography style
%   - nohyperref : do not load the hyperref package
%   - preprint : remove JMLR specific information from the template,
%         useful for example for posting to preprint servers.
%
% Example of using the package with custom options:
%
% \usepackage[abbrvbib, preprint]{jmlr2e}

\usepackage{jmlr2e}
% extra packages
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{multirow}
\usepackage{hyperref}
\hypersetup{hidelinks}
\usepackage{geometry}
\usepackage{pdflscape}
\usepackage{bm}
\usepackage{algorithm, algcompatible, algpseudocode}
\usepackage{eqparbox}

@@

% Heading arguments are {volume}{year}{pages}{date submitted}{date published}{paper id}{author-full-names}

\jmlrheading{1}{2000}{1-48}{4/00}{10/00}{meila00a}{Marina Meil\u{a} and Michael I. Jordan}

% Short headings should be running head and authors last names

\ShortHeadings{Accelerated oblique random survival forests}{Jaeger et al}
\firstpageno{1}

@@

\title{Accelerated oblique random survival forests}

\author{\name Byron C. Jaeger \email bjaeger@wakehealth.edu \\
       \addr Department of Biostatistics and Data Science\\
       Wake Forest University School of Medicine\\
       Winston-Salem, NC 27157, USA
       \AND
       \name Sawyer Welden \email swelden@wakehealth.edu \\
       \addr Department of Biostatistics and Data Science\\
       Wake Forest University School of Medicine\\
       Winston-Salem, NC 27157, USA
       \AND
       \name Kristin Lenoir \email klenoir@wakehealth.edu \\
       \addr Department of Biostatistics and Data Science\\
       Wake Forest University School of Medicine\\
       Winston-Salem, NC 27157, USA
       \AND
       \name Jaime L Speiser \email jspeiser@wakehealth.edu \\
       \addr Department of Biostatistics and Data Science\\
       Wake Forest University School of Medicine\\
       Winston-Salem, NC 27157, USA
       \AND
       \name Matthew Segar \email Matthew.Segar@UTSouthwestern.edu \\
       \addr Division of Cardiology, Department of Internal Medicine, \\
       University of Texas Southwestern Medical Center, Dallas
       \AND
       \name Nicholas M. Pajewski \email npajewsk@wakehealth.edu \\
       \addr Department of Biostatistics and Data Science\\
       Wake Forest University School of Medicine\\
       Winston-Salem, NC 27157, USA}

\editor{TBD}

@@

bm_pred_n_runs <- max(bm_pred_clean$run)

bm_pred_total_time <-
  sum(with(bm_pred_clean, time_fit + time_pred)) / (60^2)

@@

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file

The oblique random survival forest (RSF) is an ensemble supervised learning method for right-censored outcomes. Trees in the oblique RSF are grown using linear combinations of predictors to create branches, whereas in the standard RSF a single predictor is used. Oblique RSF ensembles often have higher prediction accuracy than standard RSF ensembles, but the additional computational overhead of assessing all possible linear combination of predictors is a severe limitation. In addition, few methods have been developed for interpretation of oblique RSF ensembles, and they remain something of a black box compared to their axis-based counterparts. In this article, we introduce and evaluate a method to increase computational efficiency of the oblique RSF and a method to estimate importance of individual predictor variables with the oblique RSF. Our strategy to reduce computational overhead makes use of Newton-Raphson scoring, a classical optimization technique that we apply to the Cox partial likelihood function within each non-leaf node of decision trees. We estimate importance of predictors for the oblique RSF by negating each coefficient used for the given predictor in linear combinations, and then computing the reduction in out-of-bag accuracy. In numeric experiments, we find that our implementation of the oblique RSF is over 500 times faster with equivalent discrimination and superior Brier score compared to existing software for oblique RSFs. We find in simulation studies that `negation importance' discriminates between signal and noise predictors more reliably than permutation importance, Shapley additive explanations, and a previously introduced technique to measure variable importance with oblique RSFs based on analysis of variance. All methods pertaining to oblique RSFs in the current study are available in the \texttt{aorsf} R package.

@@

\begin{keywords}
  Random Forests, Survival, Efficient, Variable Importance
\end{keywords}

@@

Oblique trees have at least two notable drawbacks compared to axis-based trees. First, finding a locally optimal oblique decision rule may require exponentially more computation than an axis-based rule. If $p$ predictors are potentially used to split $n$ observations, up to $\mathcal{O}(n^p)$ oblique splits can be assessed versus $\mathcal{O}(n \cdot p)$ axis-based splits \citep{heath1993induction, murthy1994system}. Second, estimating variable importance (VI) using permutation (a standard method for RFs) may be less effective in ensembles of oblique trees, as permuting the values of one predictor may not destabilize decisions that are based on many predictors. Although VI is one of the most widely used strategies to interpret random forests \citep{ishwaran2019standard}, few studies have investigated VI for oblique random forests \citep[see][Section~5]{menze2011oblique}, and fewer have investigated VI specifically for the oblique RSF.

This study makes two contributions to oblique RSFs. First, we reduce their computational cost (\ie, accelerate them) with a scalable algorithm to identify linear combinations of coefficients. In a general benchmark experiment including \Sexpr{n_risk_tasks} risk prediction tasks, we show that accelerated oblique RSFs are roughly 500 times faster with equivalent or superior prediction accuracy compared to existing routines to fit oblique RSFs. Second, we improve the interpretability of oblique RSFs with `negation VI', a method to estimate VI that flips the sign of coefficients in linear combinations of predictors instead of permuting predictor values. In simulation studies where VI is estimated using permutation, analysis of variance (ANOVA; see \citet{menze2011oblique}), and approximations of Shapley values, we find that negation VI improves the oblique RSF's ability to discriminate between correlated signal and noise variables. The accelerated oblique RSF and multiple methods to compute VI for oblique RSFs (permutation, ANOVA, and negation) are available in the \texttt{aorsf} R Package.

@@

\section{Related work}

@@

After \citet{breiman2001random} introduced the axis-based and oblique RF, numerous methods were developed to grow oblique RFs for classification or regression tasks \citep{menze2011oblique, zhang2014oblique, rainforth2015canonical, zhu2015reinforcement, poona2016investigating, qiu2017oblique, tomita2020sparse, katuwal2020heterogeneous}. However, oblique splitting approaches for classification or regression may not generalize to survival \citep[\eg, see][Section~4.5.1]{zhu2013tree}, and most research involving the RSF has focused on forests with axis-based trees \citep{wang2017selective}.

Building on prior research for bagging survival trees \citep{hothorn2004bagging}, \citet{hothorn2006unbiased} developed an axis-based RSF in their framework for unbiased recursive partitioning, more commonly referred to as the conditional inference forest (CIF). \citet{zhou2016random} developed a rotation forest based on the CIF and \citet{wang2017random} developed a method for extending the predictor space of the CIF. \citet{ishwaran2008random} developed an axis-based RSF with strict adherence to the rules for growing trees proposed in \citet{breiman2001random}. A similar implementation of the RSF was implemented in the \texttt{ranger} \citep{ranger}, an R package designed for high dimensional data that allows users to fit extremely randomized survival forests \citep{geurts2006extremely}. \citet{jaeger2019oblique} developed the oblique RSF following the bootstrapping approach described in Breiman's original RF and incorporating early stopping rules from the CIF.

@@

\citet{breiman2001random} introduced permutation VI, defined for each predictor as the difference in a RF's estimated generalization error before versus after the predictor's values are randomly permuted. \citet{strobl2007bias} identified bias in permutation VI driven by variable selection bias and effects induced by bootstrap sampling, and proposed an unbiased permutation VI based on unbiased recursive partitioning (see \citet{hothorn2006unbiased}). \citet{menze2011oblique} introduced an approach to estimate VI for oblique RFs that computes an analysis of variance (ANOVA) table in non-leaf nodes to obtain p-values for each predictor contributing to the node. The ANOVA VI\footnote{\citet{menze2011oblique} name their method `oblique RF VI', but we use the name `ANOVA VI' in this article to avoid confusing Menze's approach with other approaches to estimate VI for oblique RFs.} is then defined for each predictor as the number of times a p-value associated with the predictor is $\leq 0.01$ while growing a forest. \citet{lundberg2017unified} introduced a method to estimate VI using SHapley Additive exPlanation (SHAP) values, which estimate the contribution of a predictor to a model's prediction for a given observation. SHAP VI is computed for each predictor by taking the mean absolute value of SHAP values for that predictor across all observations in a given set.

% Several supervised learning algorithms can develop prediction functions for right-censored time-to-event outcomes, henceforth referred to as survival outcomes. \cite{ishwaran2008random} developed the RF for survival, an extension of the RF for regression and classification developed by \citet{breiman2001random}. Fast algorithms to fit the RF for survival are available in the \texttt{randomForestSRC} R package \citep{randomForestSRC}. A similar implementation of the RF for survival can be found in the \texttt{ranger} R package \citep{ranger}, which is particularly suited for high dimensional data. The RF for survival can also be fit using unbiased recursive partitioning \citep{cif} via the \texttt{party} R package \citep{hothorn2010party}.

\section{The acceleracted oblique random survival forest}

Consider the usual framework for survival analysis with training data $$\dataset_{\text{train}} = \left\{ (T_i, \delta_i, \bm{x}_{i}) \right\}_{i=1}^{N_{\text{train}}}.$$ Here, $T_i$ is the event time if $\delta_i=1$ and last point of contact if $\delta_i=0$, and $\bm{x}_i$ is a vector of predictors values. Assuming there are no ties, let $t_1 < \, \ldots \, < t_m$ denote the $m$ unique event times in $\dataset_{\text{train}}$.

To accelerate the oblique RSF, we propose to identify linear combinations of predictor variables in non-leaf nodes by applying Newton Raphson scoring to the partial likelihood function of the Cox regression model:
\begin{equation}\label{eqn:cox-partial-likelihood}
L(\bm\beta) = \prod_{i=1}^m \frac{e^{\bm{x}_{j(i)}^T \bm\beta}}{\sum_{j \in R_i} e^{\bm{x}_j^T \bm\beta}},
\end{equation}
where $R_i$ is the set of indices, $j$, with $T_j \geq t_i$ (i.e., those still at risk at time $t_i$), and $j(i)$ is the index of the observation for which an event occurred at time $t_i$. Newton Raphson scoring is an extremely fast estimation procedure, and the \texttt{survival} package includes documentation that outlines how to efficiently program it \citep{therneau_survival_2022}. Briefly, a vector of estimated regression coefficients, $\hat{\beta}$, is updated in each step of the procedure based on its first derivative, $U(\hat{\beta})$, and second derivative, $H(\hat{\beta})$:
$$ \hat{\beta}^{k+1} =  \hat{\beta}^{k} + U(\hat{\beta} = \hat{\beta}^{k})\, H^{-1}(\hat{\beta} = \hat{\beta}^{k}) $$

For statistical inference, it is recommended to complete iterations of Newton Raphson scoring until a convergence threshold is met. However, to identify a valid linear combination of predictors, only one iteration of Newton Raphson scoring is needed. In \secref{sec:results_pred}, we formally test whether growing oblique survival trees using one iteration of Newton Raphson scoring is sufficient (\ie, provides equivalent prediction accuracy) compared to iterating until a convergence threshold is met.

Algorithm \ref{alg:aorsf} presents our approach to fitting an oblique survival tree in the accelerated oblique RSF using default values from the \texttt{aorsf} R package. Several steps are taken to reduce computational overhead. First, memory is conserved by conducting bootstrap resampling via randomly generated bootstrap weights. Weights are integer valued, with a weight of $v$ indicating an observation was sampled $v$ times. Second, early stopping is applied to the tree growing procedure if a statistical criterion is not met. In our case, the criterion is based on the magnitude of a log-rank test statistic corresponding to splitting the data at a current node. Third, instead of greedy recursive partitioning, we use a `good enough' approach. More specifically, instead of computing a log-rank test statistic for several different linear combinations of variables and proceeding with the highest scoring option, we identify an optimal cut-point for one linear combination of variables and assess whether using this combination will create a split that passes the criterion for splitting a node. If it does not pass the criterion, then another linear combination will be tested, with the maximum number of attempts set by the parameter \texttt{n\_retry}. Often a `good-enough` split can be found in just one attempt when the training set is large, which gives the accelerated oblique RSF a computational advantage in larger training sets compared to greedy partitioning.

\begin{algorithm}
    \caption{Accelerated oblique random survival tree using default parameters.} \label{alg:aorsf}
  \begin{algorithmic}[1]
    \Require Training data $\dataset_{\text{train}} = \left\{ (T_i, \delta_i, \bm{x}_{i}) \right\}_{i=1}^{N_{\text{train}}}$, $\text{mtry} = \sqrt{\text{ncol}(\bm{x}_{\text{train}})}$, $\text{n\_split} = 5$, $\text{n\_retry} = 3$, and $\text{split\_min\_stat} = 3.841459$
    \State $\mathcal{T} \gets \emptyset$
    \State $w \gets \text{sample}(\text{from} = \left\{0, \ldots, 10\right\},\,\text{size} = \text{nrow}(\bm{x}_{\text{train}}),\, \text{replace} = \text{T})$
    \State $\dataset_{\text{in-bag}} \gets \text{subset}(\dataset_{\text{train}},\,\text{rows} = \text{which}(w > 0))$
    \State $w \gets \text{subset}(w,\, \text{which}(w > 0))$
    \State $\text{node\_assignments} \gets \text{rep}(1,\,\text{times} = \text{nrow}(\bm{x}_{\text{in-bag}}))$
    \State $\text{nodes\_to\_split} \gets \{1\}$
     \Repeat
     \For{$\text{node} \in \text{nodes\_to\_split}$}
       \State $\text{n\_try} \gets 1$
       \State $\text{node\_rows} \gets \text{which}(\text{node\_assignments} \equiv \text{node})$
       \State $\text{node\_cols} \gets \text{sample}(\text{from} = \left\{1, \ldots, \text{ncol}(\bm{x})\right\},\, \text{size} = \text{mtry},\,\text{replace} = \text{F})$ \label{marker}
       \State $\dataset_{\text{node}} \gets \text{subset}(\dataset_{\text{in-bag}},\,\text{rows} = \text{node\_rows},\,\text{columns} = \text{node\_cols})$
       \State $\beta \gets \text{newt\_raph}(\dataset_{\text{node}},\, \text{weights} = \text{subset}(w, \text{node\_rows}),\, \text{max\_iter} = 1)$
       \State $\eta \gets \bm{x}_{\text{node}} \times \beta$
       \State $\mathcal{C} \gets \text{sample}(\text{from} = \text{unique}(\eta),\, \text{size} = \text{n\_split},\,\text{replace} = \text{F})$
       \State $c \gets \argmax_{c^* \in \mathcal{C}} \left\{ \text{log\_rank\_stat}(\eta, c^*) \right\}$
       \If{$\text{log\_rank\_stat}(\eta, c) \geq \text{split\_min\_stat}$}
         \State $\mathcal{T} \gets \text{add\_node}(\mathcal{T},\, \text{name} = \text{node},\, \text{beta} = \beta,\, \text{cutpoint} = c)$
         \State \LeftComment{0}{Right node logic omitted for brevity (identical to left node logic)}
         \State $\text{node\_left\_name} \gets \text{max}(\text{node\_assignments}) + 1$
         \State $\text{node\_left\_rows} \gets \text{subset}(\text{node\_rows},\,\text{which}(\eta \leq c))$
         \State $\text{subset}(\text{node\_assignments}, \text{node\_left\_rows}) \gets \text{node\_left\_name}$
         \If{$\text{is\_splittable}(\text{subset}(\text{node\_assignments}, \text{node\_left\_rows}))$}
           \State $\text{nodes\_to\_split} \gets \text{nodes\_to\_split} \cup \text{node\_left\_name}$
         \Else
           \State $\mathcal{T} \gets \text{add\_leaf}(\mathcal{T},\, \text{data} = \text{subset}(\dataset_{\text{node}},\,\text{rows} = \text{node\_left\_rows}))$
         \EndIf
       \ElsIf{$\text{n\_try} \leq \text{n\_retry}$}
         \State $\text{n\_try} \gets \text{n\_try} + 1$
         \State \Goto{marker}
       \Else
         \State $\mathcal{T} \gets \text{add\_leaf}(\mathcal{T},\, \text{data} =\dataset_{\text{node}})$
       \EndIf
       \State $\text{nodes\_to\_split} \gets \text{nodes\_to\_split} \setminus \text{node}$
     \EndFor
  \Until{$\text{nodes\_to\_split} = \emptyset$}
  \State \Return $\mathcal{T}$
  \end{algorithmic}
\end{algorithm}

\subsection{Negation variable importance}

>>>>

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{doi}

\usepackage{lipsum}

% extra packages
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{amsmath}
\usepackage{caption}
\usepackage{multirow}
\hypersetup{hidelinks}
\usepackage{geometry}
\usepackage{pdflscape}
\usepackage{bm}
\usepackage{algorithm, algcompatible, algpseudocode}
\usepackage{eqparbox}

@@

\newcommand{\nope}[1]{}

\title{Accelerated and interpretable oblique random survival forests}

\date{\today}	% Here you can change the date presented in the paper title

\author{ \href{https://orcid.org/0000-0001-7399-2299}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}Byron C. Jaeger} \\
	Department of Biostatistics and Data Science\\
	Wake Forest University School of Medicine\\
	Winston-Salem, NC 27157, USA \\
	\texttt{bjaeger@wakehealth.edu}
	\AND
   Sawyer Welden \\
   Department of Biostatistics and Data Science\\
   Wake Forest University School of Medicine\\
   Winston-Salem, NC 27157, USA \\
   \texttt{swelden@wakehealth.edu}
   \And
   Kristin Lenoir \\
   Department of Biostatistics and Data Science\\
   Wake Forest University School of Medicine\\
   Winston-Salem, NC 27157, USA \\
   \texttt{klenoir@wakehealth.edu}
	 \And
	 Jaime L.~Speiser \\
	 Department of Biostatistics and Data Science\\
	 Wake Forest University School of Medicine\\
	 Winston-Salem, NC 27157, USA \\
	 \texttt{jspeiser@wakehealth.edu}
	 \And
	 Matthew Segar \\
   Department of Cardiology \\
   Texas Heart Institute \\
   Houston, TX 77030, USA \\
	 \texttt{Matthew.Segar@BCM.edu}
   \And
	 Ambarish Pandey \\
   Division of Cardiology, Department of Internal Medicine \\
   University of Texas Southwestern Medical Center \\
   Dallas, TX 75235, USA \\
	 \texttt{Ambarish.Pandey@UTSouthwestern.edu}
	 \And
	 Nicholas M.~Pajewski \\
	 Department of Biostatistics and Data Science\\
	 Wake Forest University School of Medicine\\
	 Winston-Salem, NC 27157, USA \\
	 \texttt{npajewsk@wakehealth.edu}
	%% \AND
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
	%% \And
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
	%% \And
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
}

% Uncomment to remove the date
%\date{}

% Uncomment to override  the `A preprint' in the header
%\renewcommand{\headeright}{Technical Report}
%\renewcommand{\undertitle}{Technical Report}
\renewcommand{\shorttitle}{\textit{arXiv} Accelerated \& interpretable oblique RSFs}

%%% Add PDF metadata to help others organize their library
%%% Once the PDF is generated, you can check the metadata with
%%% $ pdfinfo template.pdf
\hypersetup{
pdftitle={Accelerated and interpretable oblique random survival forests},
pdfsubject={q-bio.NC, q-bio.QM},
pdfauthor={Byron C.~Jaeger},
pdfkeywords={Random Forest, Survival, Efficiency, Variable Importance},
}

@@

@@

bm_pred_n_runs <- max(bm_pred_clean$data$run)

bm_pred_total_time <-
  sum(with(bm_pred_clean$data, time_fit + time_pred)) / (60^2)

@@

\begin{abstract}

	The oblique random survival forest (RSF) is an ensemble supervised learning method for right-censored outcomes. Trees in the oblique RSF are grown using linear combinations of predictors to create branches, whereas in the standard RSF, a single predictor is used. Oblique RSF ensembles often have higher prediction accuracy than standard RSF ensembles. However, assessing all possible linear combinations of predictors induces significant computational overhead that limits applications to large-scale data sets. In addition, few methods have been developed for interpretation of oblique RSF ensembles, and they remain more difficult to interpret compared to their axis-based counterparts. In this article, we introduce and evaluate a method to increase computational efficiency of the oblique RSF and a method to estimate importance of individual predictor variables with the oblique RSF. Our strategy to reduce computational overhead makes use of Newton-Raphson scoring, a classical optimization technique that we apply to the Cox partial likelihood function within each non-leaf node of decision trees. We estimate the importance of individual predictors for the oblique RSF by negating each coefficient used for the given predictor in linear combinations, and then computing the reduction in out-of-bag accuracy. In general benchmarking experiments, we find that our implementation of the oblique RSF is over 500 times faster with equivalent discrimination and superior Brier score compared to existing software for oblique RSFs. We find in simulation studies that `negation importance' discriminates between signal and noise predictors more reliably than permutation importance, Shapley additive explanations, and a previously introduced technique to measure variable importance with oblique RSFs based on analysis of variance. All methods pertaining to oblique RSFs in the current study are available in the \texttt{aorsf} R package.

@@

% keywords can be removed
\keywords{Random Forests \and Survival \and Efficiency \and Variable Importance}

@@

Oblique trees have at least two notable drawbacks compared to axis-based trees. First, finding a locally optimal oblique decision rule may require exponentially more computation than an axis-based rule. If $p$ predictors are potentially used to split $n$ observations, up to $\mathcal{O}(n^p)$ oblique splits can be assessed versus $\mathcal{O}(n \cdot p)$ axis-based splits \citep{heath1993induction, murthy1994system}. Second, estimating variable importance (VI) using permutation (a standard method for RFs) may be less effective in ensembles of oblique trees, as permuting the values of one predictor may not destabilize decisions that are based on linear combinations of predictors. Although VI is one of the most widely used strategies to interpret random forests \citep{ishwaran2019standard}, few studies have investigated VI for oblique random forests \citep[see][Section~5]{menze2011oblique}, and fewer have investigated VI specifically for the oblique RSF.

The rest of this article is organized as follows. \secref{sec:background} reviews prior studies that have developed methods related to those introduced in the current study. In \secref{sec:aorsf}, we reduce the computational cost of oblique RSFs (\ie, accelerate them) with a scalable algorithm to identify linear combinations of coefficients. In \secref{sec:negation_vi}, we improve the interpretability of oblique RSFs with `negation VI', a method to estimate VI that flips the sign of coefficients in linear combinations of predictors instead of permuting predictor values. We evaluate these methods with general benchmarking experiments and simulation studies in \secref{sec:numeric}. In \secref{sec:discussion}, we summarize results from the current study and present ideas connecting the current work to existing frameworks and methods for RSFs that future studies may engage with. The accelerated oblique RSF and multiple methods to compute VI for oblique RSFs are available in the \texttt{aorsf} R Package.

@@

\section{Related work} \label{sec:background}

@@

After \citet{breiman2001random} introduced the axis-based and oblique RF, numerous methods were developed to grow oblique RFs for classification or regression tasks \citep{menze2011oblique, zhang2014oblique, rainforth2015canonical, zhu2015reinforcement, poona2016investigating, qiu2017oblique, tomita2020sparse, katuwal2020heterogeneous}. However, oblique splitting approaches for classification or regression may not generalize to censored outcomes \citep[\eg, see][Section~4.5.1]{zhu2013tree}, and most research involving the RSF has focused on forests with axis-based trees \citep{wang2017selective}.

Building on prior research for bagging survival trees \citep{hothorn2004bagging}, \citet{hothorn2006unbiased} developed an axis-based RSF in their framework for unbiased recursive partitioning, more commonly referred to as the conditional inference forest (CIF). \citet{zhou2016random} developed a rotation forest based on the CIF and \citet{wang2017random} developed a method for extending the predictor space of the CIF. \citet{ishwaran2008random} developed an axis-based RSF with strict adherence to the rules for growing trees proposed in \citet{breiman2001random}.  \citet{jaeger2019oblique} developed the oblique RSF following the bootstrapping approach described in Breiman's original RF and incorporating early stopping rules from the CIF.

Fast algorithms to fit axis-based RSFs are available in the \texttt{randomForestSRC} R package \citep{randomForestSRC} and the \texttt{ranger} \citep{ranger} R package. \texttt{randomForestSRC} provides a unified interface to grow RFs in a wide range of analyses, and \texttt{ranger} is designed to grow RFs efficiently using high dimensional data. Fast algorithms to fit the CIF are provided by the \texttt{party} R package \citep{hothorn2010party}, which provides a computational toolbox for recursive partitioning using conditional inference trees. \citet{jaeger2019oblique} developed the \texttt{obliqueRSF} package and found it was approximately 30 times slower than \texttt{party} and nearly 200 times slower than \texttt{randomForestSRC}. Few studies have developed software with fast algorithms for oblique RSFs that have comparable speed compared to algorithms for axis-based RSFs.

@@

Several techniques to estimate VI have been developed since \citet{breiman2001random} introduced permutation VI, which is defined for each predictor as the difference in a RF's estimated prediction error before versus after the predictor's values are randomly permuted. \citet{strobl2007bias} identified bias in permutation VI driven by variable selection bias and effects induced by bootstrap sampling, and proposed an unbiased permutation VI measure based on unbiased recursive partitioning \citep{hothorn2006unbiased}. \citet{menze2011oblique} introduced an approach to estimate VI for oblique RFs that computes an analysis of variance (ANOVA) table in non-leaf nodes to obtain p-values for each predictor contributing to the node. The ANOVA VI\footnote{\citet{menze2011oblique} name their method `oblique RF VI', but we use the name `ANOVA VI' in this article to avoid confusing Menze's approach with other approaches to estimate VI for oblique RFs.} is then defined for each predictor as the number of times a p-value associated with the predictor is $\leq 0.01$ while growing a forest. \citet{lundberg2017unified} introduced a method to estimate VI using SHapley Additive exPlanation (SHAP) values, which estimate the contribution of a predictor to a model's prediction for a given observation. SHAP VI is computed for each predictor by taking the mean absolute value of SHAP values for that predictor across all observations in a given set. With the exception of \citet{menze2011oblique}, few studies have evaluated estimation of VI using oblique RFs, and fewer have examined VI specifically for the oblique RSF.

% Several supervised learning algorithms can develop prediction functions for right-censored time-to-event outcomes, henceforth referred to as survival outcomes. \cite{ishwaran2008random} developed the RF for survival, an extension of the RF for regression and classification developed by \citet{breiman2001random}.

\section{The accelerated oblique random survival forest} \label{sec:aorsf}

This section describes our approach to reduce computational overhead of the oblique RSF. Consider the usual framework for right-censored time-to-event outcomes with training data $$\dataset_{\text{train}} = \left\{ (T_i, \delta_i, \bm{x}_{i}) \right\}_{i=1}^{N_{\text{train}}}.$$ Here, $T_i$ is the event time if $\delta_i=1$ or the censoring time if $\delta_i=0$, and $\bm{x}_i$ is a vector of predictors values. Assuming there are no ties, let $t_1 < \, \ldots \, < t_m$ denote the $m$ unique event times in $\dataset_{\text{train}}$.

To accelerate the oblique RSF, we propose to identify linear combinations of predictor variables in non-leaf nodes by applying Newton Raphson scoring to the partial likelihood function of the Cox regression model:
\begin{equation}\label{eqn:cox-partial-likelihood}
L(\bm\beta) = \prod_{i=1}^m \frac{e^{\bm{x}_{j(i)}^T \bm\beta}}{\sum_{j \in R_i} e^{\bm{x}_j^T \bm\beta}},
\end{equation}
where $R_i$ is the set of indices, $j$, with $T_j \geq t_i$ (i.e., those still at risk at time $t_i$), and $j(i)$ is the index of the observation for which an event occurred at time $t_i$. Newton Raphson scoring is an exceptionally fast estimation procedure, and the \texttt{survival} package includes documentation that outlines how to efficiently program it \citep{therneau_survival_2022}. Briefly, a vector of estimated regression coefficients, $\hat{\beta}$, is updated in each step of the procedure based on its first derivative, $U(\hat{\beta})$, and second derivative, $H(\hat{\beta})$:
$$ \hat{\beta}^{k+1} =  \hat{\beta}^{k} + U(\hat{\beta} = \hat{\beta}^{k})\, H^{-1}(\hat{\beta} = \hat{\beta}^{k}) $$

For statistical inference, it is recommended to continue updating $\hat{\beta}$ by completing additional iterations of Newton Raphson scoring until a convergence threshold is met. However, since an estimate of $\hat{\beta}$ is created by the first iteration of Newton Raphson scoring, only one iteration of Newton Raphson scoring is needed to identify a valid linear combination of predictors. Moreover, computing $U$ and $H$ requires computation and exponentiation of the vector $\bm{x}\hat{\beta}$, but these steps can be skipped on the first iteration of Newton Raphson scoring if an initial value of $\hat{\beta} = 0$ is chosen, allowing for a reduction in computing operations and removing the need to scale predictor values prior to initiating the Newton Raphson algorithm.\footnote{Predictors are scaled prior to initiating the Newton Raphson algorithm to avoid exponentiation of large numbers. However, if only one iteration is completed with an initial value of 0 for $\hat{\beta}$, then $\exp(\bm{x}\hat{\beta}) = 1$.} In \secref{sec:results_pred}, we formally test whether growing oblique survival trees using one iteration of Newton Raphson scoring provides equivalent prediction accuracy compared to trees where iterations are completed until a convergence threshold is met.

Algorithm \ref{alg:aorsf} presents our approach to fitting an oblique survival tree in the accelerated oblique RSF using default values from the \texttt{aorsf} R package. Several steps are taken to reduce computational overhead. First, memory is conserved by conducting bootstrap resampling via randomly generated bootstrap weights rather than making a traditional bootstrap sample. Weights are integer valued, with a weight of $v$ indicating an observation was sampled $v$ times. Second, early stopping is applied to the tree-growing procedure if a statistical criterion is not met. In our case, the criterion is based on the magnitude of a log-rank test statistic corresponding to splitting the data at a current node. Third, instead of greedy recursive partitioning, we use `good enough' partitioning. More specifically, instead of computing a log-rank test statistic for several different linear combinations of variables and proceeding with the highest scoring option, we identify an optimal cut-point for one linear combination of variables and assess whether using this combination will create a split that passes the criterion for splitting a node. If it does not pass the criterion, then another linear combination will be tested, with the maximum number of attempts set by the parameter \texttt{n\_retry}. Often a `good-enough` split can be found in just one attempt when the training set is large, which gives the accelerated oblique RSF a computational advantage in larger training sets compared to greedy partitioning.
